Demystifying the Computer Science \\Ph.D. Admission in US Universities\\{\large A guide for Vietnamese and International Students

### 2023-06-12 phd-cs-us.tex
updates

\newcommand{\mycomment}[3][\color{black}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\tl}[1]{\mycomment{ThanhLe}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\title{Demystifying the Computer Science \\Ph.D. Admission in US Universities\\{\large A guide for Vietnamese and International Students}}
\date{\today}
\author{\small \href{https://nguyenthanhvuh.github.io}{ThanhVu (Vu) Nguyen}, \href{https://cs.gmu.edu}{George Mason University}}
%\setcounter{secnumdepth}{3}
\begin{document}
\maketitle

@@

\begin{enumerate}   
    \item What can you do to increase your admission chance?
      \begin{itemize}[leftmargin=*]
        \item Show something that makes you \textbf{stand out}, e.g., are you a female or a minority in CS (research for this on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS?  All of these are important in CS in the US.

    \item Even if you do not have formal research experience, you can talk about your personal projects. For example, if you have an open-source project on Github that is used by many people, has lots of stars in Github, do talk about it. If you write technical, research-like blogs, talk about them too.
    \end{itemize}

   \item How to rank or select a CS program?

   \begin{center}
    \includegraphics[scale=0.5]{c1.png}
   \end{center}

\begin{itemize}[leftmargin=*]
\item  International students not familiar with US universities often put them into \emph{two} bins:  (i) very top school that they dream about such as Stanford, MIT, Princeton, Harvard and (ii) everything else.  Sometimes they rank CS programs using the reputations non-CS programs, e.g., medical or physics.
In some cases they rank universities based on popular states they know in the US, e.g., California, New York, and Texas.  Let's just say there are so many thing wrong with these methods.

\item You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}, which is designed specifically to help prospective PhD students in Computer Science!  You will be very surprised to learn that a school X that you don't know much about has very strong research in your interested field Y. This is also a good way to learn about individual faculty (who works on what) and the well-known CS conferences\footnote{In CS (and only in CS), conferences, not journals, are oten the main venue to publish research finding.}.
\begin{commentbox}
    Dat: many Vietnamese students, even those from very top engineering schools, do not know about CSRankings.  Only few applicants who worked at top research places such as VinAI might know about it.
\end{commentbox}

\item More generally, rankings can be superficial and you will want to do more research to be informed and make better decision. For example, if you get admissions to several places, you should contact profs. that you're interested in at those place and talk to them. They would be more willing to chat to you now that you have been admitted.  Ask them questions about their work, how they manage students, their expectations. You can even ask to contact their students.
\end{itemize}

    \item Can I apply to PhD in CS if my undergrad was not in CS or any related areas?
    \begin{itemize}[leftmargin=*]
        \item Yes, as long as you can demonstrate you are ready for CS PhD research through research experiences, LoRs, statements, etc as mentioned. You might be even able leverage this to make your profile stand out.
    \end{itemize}

    \item   Is an MS degree required for admission to PhD?
    \begin{itemize}[leftmargin=*]
    \item No. In fact, student with BS can get MS degree "along the way" to PhD.  However, MS can help if it gives research experience or is from a more well-known school than your undergrad institution.
    \end{itemize}

    \item How do I address a professor Firstname Lastname?
    \begin{itemize}[leftmargin=*]
        \item If you don't know that professor (e.g., first email contact), then use \textbf{Prof. Lastname} or \textbf{Dr. Lastname}. I've seen many international students write \textbf{Prof.} or \textbf{Dr.} \textbf{Firstname Lastname}.  Writing like that makes it like you copy and paste the names, so no need to do so,  just Prof. or Dr. Lastname.

 \item If you don't know the prof., do not use Mr. or Mrs., or Firstname. To me it seems a bit disrespectful. As you know that prof. better and depends on their preference, you can call them by their Firstname.

\begin{commentbox}
    Vu: I've been called Dr. Vu and I find it a bit amusing but am totally fine with it.
    \end{commentbox}
    \end{itemize}
\end{enumerate}

>>>>

\newcommand{\mycomment}[3][\color{black}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\tl}[1]{\mycomment{ThanhLe}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\title{Demystifying the Computer Science \\Ph.D. Admission in US Universities\\{\large A guide for Vietnamese and International Students}}
%\date{\today}
\author{\small \href{https://nguyenthanhvuh.github.io}{ThanhVu (Vu) Nguyen}\\
\href{https://cs.gmu.edu}{George Mason University}}
%\setcounter{secnumdepth}{3}
\begin{document}
\maketitle

@@

\subsection{What can you do to increase your admission chance?}
 Show something that makes you \textbf{stand out}, e.g., are you a female or a minority in CS (research for this on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS?  All of these are important in CS in the US.

Even if you do not have research experience, you can talk about your personal projects. For example, if you have an open-source project on Github that is used by many people, has lots of stars in Github, do talk about it. If you write technical, research-like blogs, talk about them too.

\subsection{How to rank or select a CS program?}
   \begin{center}
    \includegraphics[scale=0.5]{c1.png}
   \end{center}

International students not familiar with US universities often put them into \emph{two} bins:  (i) very top school that they dream about such as Stanford, MIT, Princeton, Harvard and (ii) everything else.  Sometimes they rank CS programs using the reputations non-CS programs, e.g., medical or physics.
In some cases they rank universities based on popular states they know in the US, e.g., California and New York.  Let's just say there are so many thing wrong with these methods.

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}, which is designed specifically to help prospective PhD students in Computer Science!  You will be very surprised to learn that a school X that you don't know much about has very strong research in your interested field Y. This is also a good way to learn about individual faculty (who works on what) and the well-known CS conferences\footnote{In CS (and only in CS), conferences, not journals, are oten the main venue to publish research finding.}.
\begin{commentbox}
    Dat: many Vietnamese students, including those from very top engineering schools, do not know about CSRankings.  May be applicants who worked at top research places such as VinAI might know about it.
\end{commentbox}

More generally, rankings can be superficial and you will want to do more research to be informed and make better decision. For example, if you get admissions to several places, you should contact profs. that you're interested in at those place and talk to them. They would be more willing to chat to you now that you have been admitted.  Ask them questions about their work, how they manage students, their expectations. You can even ask to contact their students.

\subsection{Can I apply to PhD in CS if my undergrad was not in CS or any related areas?}

Yes, as long as you can demonstrate you are ready for CS PhD research through research experiences, LoRs, statements, etc as mentioned. You might be even able leverage this to make your profile stand out.

\subsection{Is an MS degree required for admission to PhD?}
No. In fact, student with BS can get MS degree "along the way" to PhD.  However, MS can help if it gives research experience or is from a more well-known school than your undergrad institution.

\subsection{How do I address a professor Firstname Lastname?}
If you don't know that professor (e.g., first email contact), then use \textbf{Prof. Lastname} or \textbf{Dr. Lastname}. I've seen many international students write \textbf{Prof.} or \textbf{Dr.} \textbf{Firstname Lastname}.  Writing like that makes it like you copy and paste the names, so no need to do so,  just Prof. or Dr. Lastname.

If you don't know the prof., do not use Mr. or Mrs., or Firstname. To me it seems a bit disrespectful. As you know that prof. better and depends on their preference, you can call them by their Firstname.

\begin{commentbox}
    Vu: I've been called Dr. Vu and I find it a bit amusing but am totally fine with it.
\end{commentbox}
